---
title: "DATA608_Project Analysis"
author: "Chinedu Onyeka and Charles Ugiagbe"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Load Libraries**
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(readr)
library(gridExtra)
library(scales)
library(lubridate)
library(ggrepel)
library(rgdal)
library(plotly)
library(kableExtra)
library(shiny)
```


**Read the data**
```{r read-data, message=FALSE, warning=FALSE}
url_data <- "https://raw.githubusercontent.com/chinedu2301/data608-project/main/data/motor_crash_df.csv"
crash_df <- read.csv(url_data)
crash_df <- as_tibble(crash_df) 
```


**Change the crash_date data type to date format and View the head of the data**
```{r}
crash_df$crash_date <- as.Date(crash_df$crash_date, format = "%Y-%m-%d")
crash_df %>% head(8)
```


**Take a glimpse of the data**
```{r}
glimpse(crash_df)
```

###   Car crashes by year

*Since 2018, there has been a decline in the car crash incident in NYC but there was a sharp decline from 2019 to 2020. The decline has also continue steadily after then. The 2020 sharp decline might be due to covid-related lockdowns. Also, increase in remote work opportunities may be a reason for decline in incidents, but more studies need to be conducted to ascertain that.*

```{r}
crash_df$year <- year(crash_df$crash_date) #extract year from date using the lubridate year() function

crash_df %>%
        ggplot(aes(x=as.factor(year))) + geom_bar(stat='count', fill='purple') +
        scale_y_continuous(labels=comma) + labs(x='Year', y='No of Incident', title='Incidents by year') + 
        geom_label(stat = "count", aes(label = ..count.., y = ..count..)) + 
        theme(axis.title = element_text(size = 13), plot.title = element_text(size = 15,
        hjust = 0.5), panel.background = element_rect(fill = "lemonchiffon")) +
        labs(title = "Incidents by Year", y = "No of Incidents")
```

### Car crash by quarter

```{r}
library(viridis)
crash_df$quarter <- quarter(crash_df$crash_date) #extract Quarters from date

q1 <- crash_df %>% filter(year!=2013) %>% select(year, quarter) %>% group_by(year) %>% count(quarter) %>%
        ggplot(aes(x=as.factor(quarter), y=n, fill = quarter)) + 
        geom_bar(stat='identity') + scale_y_continuous(labels=comma) + facet_grid(.~year) + 
        labs(x='Quarter', y='Number of incidents', title='Incidents by Quarter') 

q1 + scale_fill_viridis() + theme(axis.title = element_text(size = 13),
      plot.title = element_text(size = 15,
      hjust = 0.5), panel.background = element_rect(fill = "lemonchiffon"))
```


###   Car crashes by weekdays

*As expected, Friday has the most crash incidents because it is usually the most busy day in terms of vehicular movement. People travelling for weekend.*
```{r}
crash_df$weekday <- wday(crash_df$crash_date, label=TRUE)

crash_df %>% count(weekday) %>%
        ggplot(aes(x=weekday, y=n)) + geom_bar(stat='identity', fill=rainbow(n=7)) +
        scale_y_continuous(labels=comma) +
        labs(x='Weekday', y='Number of incidents', title='Incidents by Weekday') + 
        theme(axis.title = element_text(size = 13),
        plot.title = element_text(size = 15,
        hjust = 0.5), panel.background = element_rect(fill = "lemonchiffon"))
```

### Crash incidents by Time of Day
 
```{r}
# Extract time of day from the data
time <- as.POSIXct(strptime(c(crash_df$crash_time), "%H:%M"), "UTC")
x=as.POSIXct(strptime(c("0000","0500","1100","1600","1900","2359"),
                      "%H%M"),"UTC")
labs=c("night","morning","afternoon","evening","night")
day_time <- labs[findInterval(time,x)]

crash_df$day_time <- c(day_time)
print(crash_df)
```

```{r}
day_time_df <- crash_df %>% group_by(day_time) %>% summarise(count_incidents = n()) 
day_time_df
```

```{r}
day_time_df %>% ggplot(aes(x=reorder(day_time, count_incidents), y=count_incidents)) + 
        geom_bar(stat='identity', fill='purple') +
        scale_y_continuous(labels=comma) +
        labs(x='new_time', y='Number of incidents', title='Incidents by time') + theme(axis.title = element_text(size = 13),
        plot.title = element_text(size = 15,
        hjust = 0.5), panel.background = element_rect(fill = "lemonchiffon")) +
        labs(title = "Incidents by time of day", x = "time_of_day")
```

###   Crash Incident by Borough.

*Brooklyn has the highest number of crashes and Staten Island has the least number of crash incidents.*
  
```{r}
plotly::ggplotly(crash_df %>% count(borough) %>%
        ggplot(aes(x=reorder(borough, n), y=n, fill=n, text=borough)) +
        geom_bar(stat='identity', fill='violet') +
        labs(x='', y='Number of crashes'),
        tooltip=c("text", "y"))
```

### Top Ten(10) incidents by number of victims

```{r}
crash_df$victims <- crash_df$number_killed + crash_df$number_injured
Top10 <- crash_df %>% select(crash_date, borough, number_killed, number_injured, victims, zip_code) %>% 
         arrange(desc(victims)) %>% top_n(n=10, wt=victims)
Top10
```


### Victims per crash incident
```{r}
crash_df$victims <- crash_df$number_killed + crash_df$number_injured

VictimsByborough <- crash_df %>% group_by(borough)  %>%   
                    summarize(sumVic=sum(victims), sumInj=sum(number_injured), sumDeath=sum(number_killed),
                              PercDeath=round(sumDeath/sumVic,3), sumIncidents=n(), 
                              vicPerInc=round(sumVic/sumIncidents,1)) %>% arrange(desc(sumVic))
VictimsByborough
```

```{r}
ggplotly(crash_df %>% count(contributing_factor) %>%
        ggplot(aes(x=contributing_factor, y=n, fill=n, text=contributing_factor)) +
        geom_bar(stat='identity', fill='red') + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(x='', y='Number of crashes'),
         tooltip=c("text", "y"))
```

```{r}
df <- crash_df
choices <- c("Crash by Year", "Crash by Quarter", "Crash by Weekday", "Crash by Time of Day", "Crash by Borough",
             "Victims per Incident")
```


```{r shinny1}
# Build the UI for question1
ui <- fluidPage(
  # Panel for Crash by Year
  tabsetPanel( 
    tabPanel(title = "NYC Vehicular Incidents",
      sidebarPanel(
        htmlOutput('message_q1'),
        selectInput('plotType', 'Plot to Display', 
                  unique(choices), selected= choices[0],width = 600)
      ),
      mainPanel(plotlyOutput('plot1_q1'))
    )
    ))

# build the server for question1
server <- function(input, output, session) {
  # Question1 Server Side
  data_year <- reactive({
    df2 <- df %>%
        ggplot(aes(x=as.factor(year))) + geom_bar(stat='count', fill='purple') +
        scale_y_continuous(labels=comma) + labs(x='Year', y='No of Incident', title='Incidents by year') + 
        geom_label(stat = "count", aes(label = ..count.., y = ..count..)) + 
        theme(axis.title = element_text(size = 13), plot.title = element_text(size = 15,
        hjust = 0.5), panel.background = element_rect(fill = "lemonchiffon")) +
        labs(title = "Incidents by Year", y = "No of Incidents")
    df2
  })
  
  data_quarter <- reactive({
    df2 <- df %>% filter(year!=2013) %>% select(year, quarter) %>% group_by(year) %>% count(quarter) %>%
        ggplot(aes(x=as.factor(quarter), y=n, fill = quarter)) + 
        geom_bar(stat='identity') + scale_y_continuous(labels=comma) + facet_grid(.~year) + 
        labs(x='Quarter', y='Number of incidents', title='Incidents by Quarter') + scale_fill_viridis() + 
        theme(axis.title = element_text(size = 13),
        plot.title = element_text(size = 15,
        hjust = 0.5), panel.background = element_rect(fill = "lemonchiffon"))
    
  })
  
  data_weekday <- reactive({
    df2 <- df %>% count(weekday) %>%
        ggplot(aes(x=weekday, y=n)) + geom_bar(stat='identity', fill=rainbow(n=7)) +
        scale_y_continuous(labels=comma) +
        labs(x='Weekday', y='Number of incidents', title='Incidents by Weekday') + 
        theme(axis.title = element_text(size = 13),
        plot.title = element_text(size = 15,
        hjust = 0.5), panel.background = element_rect(fill = "lemonchiffon"))
    
  })

  data_timeofday <- reactive({
    df2 <- day_time_df %>% ggplot(aes(x=reorder(day_time, count_incidents), y=count_incidents)) + 
        geom_bar(stat='identity', fill='purple') +
        scale_y_continuous(labels=comma) +
        labs(x='new_time', y='Number of incidents', title='Incidents by time') + theme(axis.title = element_text(size = 13),
        plot.title = element_text(size = 15,
        hjust = 0.5), panel.background = element_rect(fill = "lemonchiffon")) +
        labs(title = "Incidents by time of day", x = "time_of_day")
    
  })    
  
  data_borough <- reactive({
    df2 <- plotly::ggplotly(df %>% count(borough) %>%
        ggplot(aes(x=reorder(borough, n), y=n, fill=n, text=borough)) +
        geom_bar(stat='identity', fill='violet') +
        labs(x='', y='Number of crashes'),
        tooltip=c("text", "y"))
    
  })
  
  data_victims <- reactive({
    df2 <- ggplotly(df %>% count(contributing_factor) %>%
        ggplot(aes(x=contributing_factor, y=n, fill=n, text=contributing_factor)) +
        geom_bar(stat='identity', fill='red') + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(x='', y='Number of crashes'),
         tooltip=c("text", "y"))
    
  })
  


  # output Plot
  
  output$plot1_q1 <- renderPlotly({
    if (input$plotType == "Crash by Year"){
      df2 <- data_year()
    } else if(input$plotType == "Crash by Quarter"){
      df2 <- data_quarter()
    } else if(input$plotType == "Crash by Weekday"){
      df2 <- data_weekday()
    } else if(input$plotType == "Crash by Time of Day"){
      df2 <- data_timeofday()
    } else if(input$plotType == "Crash by Borough"){
      df2 <- data_borough()
    } else{
      df2 <- data_victims()
    }
  }
  
  )
}

shinyApp(ui = ui, server = server)
```


