---
title: "data608-project"
author: "Chinedu Onyeka and Charles Ugiagbe"
date: '2022-11-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Load required libraries**
```{r load-lib, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(plotly)
library(shiny)
```

**Read the data**
```{r read-data,message=FALSE, warning=FALSE}
url <- "MotorVehicleCollisionsCrashes.csv"
motorCrash <- read.csv(url)
motorCrash <- as_tibble(motorCrash)
```

**Check the head of the data**
```{r}
head(motorCrash)
```
<center> <h3>  Data Pre-processing  </h3> </center>

**Select required columns and rename the columns**
```{r}
req_columns <- motorCrash %>% select("CRASH.DATE", "CRASH.TIME", "BOROUGH", "ZIP.CODE", "LATITUDE", "LONGITUDE",
                                     "NUMBER.OF.PERSONS.INJURED", "NUMBER.OF.PERSONS.KILLED", "CONTRIBUTING.FACTOR.VEHICLE.1",
                                     "VEHICLE.TYPE.CODE.1")
headers <- c("crash_date", "crash_time", "borough", "zip_code", "latitude", "longitude", "number_injured", "number_killed",
            "contributing_factor","vehicle_type")
names(req_columns) <- headers
head(req_columns)
```
**Remove records that have missing zip_code, latitude, longitude, vehicle_type, and contributing_factor**
```{r}
missing <- c(NA, NULL, "")
remove_missing <- req_columns %>% filter(!zip_code %in% missing & !latitude %in% missing & !longitude %in% missing &
                                           !vehicle_type %in% missing & !contributing_factor %in% missing)
head(remove_missing, 20)
```
**Convert the crash_date column to date format**
```{r}
remove_missing$crash_date <- as.Date(remove_missing$crash_date, format = "%m/%d/%Y")
```

**Filter the data for only records with year greater than 2015**
```{r}
motor_crash = remove_missing %>% filter(year(crash_date) > 2015)
head(motor_crash)
```

**Drop the remaining NA**
```{r}
motor_crash_df <- motor_crash %>% drop_na()
```

**Check if there is any NA in the dataset**
```{r}
any(is.na(motor_crash_df))
```
**Take a glimpse of the data**
```{r}
glimpse(motor_crash_df)
```

**Look at the summary statistics of the data**
```{r}
summary(motor_crash_df)
```


```{r}
file = "C:\\Users\\exper\\OneDrive\\Documents\\CUNY SPS\\Fall 2022\\DATA_608\\project\\data608-project\\data608-project\\data\\motor_crash_df.csv"
write_csv(motor_crash_df, file = file)
```

<center> <h3>  Data Analysis  </h3> </center>

**Read the data from github**
```{r read-data-github,message=FALSE, warning=FALSE}
url_data <- "https://raw.githubusercontent.com/chinedu2301/data608-project/main/data/motor_crash_df.csv"
crash_df <- read.csv(url_data)
crash_df <- as_tibble(crash_df) 
```


**Change the crash_date data type to date format and View the head of the data**
```{r}
crash_df$crash_date <- as.Date(crash_df$crash_date, format = "%Y-%m-%d")
crash_df %>% head(20)
```


****
```{r}
###
```








